Omics-based Single Step Trait Prediction

### 2017-09-14 paper/manuscript.tex
paper, clarify predictions

% Please add the following required packages to your document preamble:
% \usepackage{booktabs}
\begin{table}[]
\centering
\caption{Acronyms for the trait names}
\label{table:TraitKey}
\begin{tabular}{@{}lll@{}}
\toprule
Trait              & Abbreviation & Data Set  & Experiment\\ \midrule
Cob weight         & CW           & Diversity & Exp1 \\
Days to silking    & DS           & Diversity & Exp1 \\
Ear diameter       & ED           & Diversity & Exp1 \\
100-grain weight   & GW           & Diversity & Exp1 \\
Kernel width       & KW           & Diversity & Exp1 \\
Plant height       & PH           & Diversity & Exp1 \\
Dry matter yield   & DMY          & Hybrid    & Exp2 \\
Dry matter content & DMC          & Hybrid    & Exp2 \\
Lignin             & ADL          & Hybrid    & Exp2 \\
Fat                & FAT          & Hybrid    & Exp2 \\
Protein            & PRO          & Hybrid    & Exp2 \\
Starch             & STA          & Hybrid    & Exp2 \\
Sugar              & SUG          & Hybrid    & Exp2 \\ \bottomrule
\end{tabular}
\end{table}

@@

\Genetics2level{Prediction}
\subsection{Kernels}
BLUP models are computationally attractive when the number of features $p$
exceeds the number of genotypes $n$ and are equivalent to a selection index
when fixed effects have been accomodated by the dependent variable 
\cite{Mrode2014}.
Depending on the data set, up to three predictors were available for agronomic
trait predictions, namely pedigree data (P), genomic data (G) and
transcriptompic data (T).
The corresponding feature matrix of the $l$-th group of inbred lines 
($\mathbf{W}_{l}$) has dimensions $n_{l} \times p$ where $n_{l}$ pertains to 
the number of genotypes in the $l$-th group and $p$ pertains to the number of 
features.
Note that $l$ is only defined for Exp2 where a distinction between 
the two heterotic groups is necessary whereas $l$ can be ignored for all 
models pertaining to the inbred lines of the maize diversity panel in 
Exp1.
For the diversity panel maize lines, $n_{l}$ corresponds to the number of 
genotypes whereas, in the case of the hybrid data, $n_{l}$ corresponds to the 
number of parent lines in the $l$-th heterotic group.
In generating kernels for each predictor, all features in $\mathbf{W}_{l}$ were
centered and standardized to unit variance.
Then, each kernel can be defined as

\begin{equation} \label{eq:GenomicRelationship}
  \mathbf{K}_{l} = \frac{1}{W_{l}} \mathbf{W}_{l} \mathbf{W}_{l}^{\top},
\end{equation}

where $W_{l}$ denotes the number of features~\cite{VanRaden2008}.
In the case of pedigree data (P), coancestry coefficients were used directly
for $\mathbf{K}_{l}$.

\subsection{Breeding value model}
The general model for breeding values was:

\begin{equation} \label{eq:KBLUPModel}
  \mathbf{y} = \mu + 
  \sum_{l=1}^{L} \mathbf{Z}_{l} \mathbf{g}_{l} + \mathbf{e},
\end{equation}

where $\mathbf{y}$ is the vector of observed inbred line or hybrid performance,
respectively, $\mu$ is the fixed model intercept, and $\mathbf{e}$ is the 
residual error.
For the hybrid data from Exp2, we fit separate additive effects for 
each of the $L$ heterotic groups and $\mathbf{g}_{l}$ now pertains to the 
general combining ability effects of the parental inbred lines (either Dent or 
Flint, as expressed by $l$).
For the maize diversity panel inbred lines from Exp1, we fit random 
genotype effects $\mathbf{g}_{l}$ (\textit{i.e.}, breeding values).
For both experiments, genetic effects $\mathbf{g}_{l}$ are associated with 
the vector of phenotypic performance $\mathbf{y}$ through a corresponding 
design matrix $\mathbf{Z}_{l}$.
The random effects ($\mathbf{g}_{l}$) have expectation zero and variance equal 
to $\mathbf{K}_{l} \sigma^{2}_{{g}^{l}}$ and $\mathbf{I} \sigma^2_{e}$ 
for the residual error.
Here, we did not consider the inclusion of dominance effects for the hybrid
material after a previous investigation into the predictive ability of various
``omics'' predictors in the same material concluded no considerably gains 
\cite{Westhues2017}.
All prediction models were implemented using the R-packages \textit{BGLR}
\cite{Perez2014} and \textit{sspredr}
\url{https://github.com/mwesthues/sspredr}.

\subsection{Single step prediction}
For simplicity, let us assume throughout this section that we have $L=1$.
If the breeding values are fully captured by the features in $\mathbf{W}$,
Eq.~\ref{eq:KBLUPModel} can then be expressed as

\begin{align} \label{eq:KBLUPModelSimplified}
	\mathbf{y} &= \mu + \mathbf{Z} \mathbf{g} + \boldsymbol{e} \\
	&= \mu + \mathbf{Z} \mathbf{W} \boldsymbol{\alpha} + \boldsymbol{e}
\end{align}
\cite{Fernando2014} since the ridge regression model is equivalent to the BLUP
model~\cite{Ruppert2003}.

Hence, the breeding values in a prediction model are given by:
\begin{equation} \label{eq:mrnaebv}
	\mathbf{\hat{g}} = \mathbf{W}\boldsymbol{\hat{\alpha}},
\end{equation}

where $\boldsymbol{\hat{\alpha}}$ is the solution to a ridge regression model.

Consider now the situation in which one predictor is complete in the sense that
the whole population is equipped with features for that predictor whereas only
a subset of that population has information on a second predictor.
We can take a similar route as in \citeNP{Fernando2014} and impute covariates 
of the incomplete predictor by using covariates of the complete predictor.
Let the subscript $1$ denote individuals covered only by the complete
predictor whereas individuals covered by both, the complete and the
incomplete predictor, are indicated by the subscript $2$.
Further, the covariates in $\mathbf{W}$ are centered.
In this case we need to partition the vector of breeding values $\mathbf{g}$
into a component $\mathbf{g}_{1}$ (\textit{i.e.,} breeding values of animals 
covered only by the complete predictor) and a component 
$\mathbf{g}_{2} = \mathbf{W}_{2}\boldsymbol{\alpha}$ (\textit{i.e.,} breeding 
values of animals covered by both predictors).
Under multivariate normality, the vector $\mathbf{g}_1$ can be written as the
sum of the conditional expectation given $\mathbf{g}_2$ and a residual error
term:
\begin{align} \label{eq:mrna1}
	\mathbf{g_1} &= E(\mathbf{g}_1|\mathbf{g}_2) + \boldsymbol{\epsilon} \\
	&= \mathbf{K_{12}}\mathbf{K_{22}}^{-1}\mathbf{W_2}\boldsymbol{\hat{\alpha}} + (\mathbf{g_1} - \mathbf{K_{12}}\mathbf{K_{22}}^{-1}\mathbf{W_2}\boldsymbol{\hat{\alpha}}) \\
	&= \mathbf{\hat{g}}_1 + \boldsymbol{\epsilon},
\end{align}

where $\mathbf{K}_{ij}$ are partitions of $\mathbf{K}$ that correspond to 
$\mathbf{g}_{1}$ and $\mathbf{g}_{2}$, respectively~\cite{Fernando2014}.

The covariance matrix of $\boldsymbol{\epsilon}$ is $\mathbf{K}_{11} - \mathbf{K}_{12}\mathbf{K}_{22}^{-1}\mathbf{K}_{21} = (\mathbf{K}^{11})^{-1}$
\cite{Legarra2009}.

@@

\begin{equation} \label{eq:single-step-submatrices}
\mathbf{X} = 
\begin{bmatrix}
  X_1 \\
  X_2 
 \end{bmatrix},
 \mathbf{W} = 
\begin{bmatrix}
  Z_1\hat{\mathbf{W}_1} \\
  \mathbf{W}_2 
 \end{bmatrix},
 \mathbf{U} = 
\begin{bmatrix}
  Z_1 \\
  0 
 \end{bmatrix},
\end{equation}

where $\hat{\mathbf{W}}_{1} = \mathbf{K}_{12} \mathbf{K}_{22}^{-1} \mathbf{W}_{2}$.

@@

\begin{align} \label{eq:entangled-augmented-single-step-model}
\begin{bmatrix}
  y_1 \\
  y_2 
 \end{bmatrix}
& =
 \begin{bmatrix}
  X_1 \\
  X_2 
 \end{bmatrix}
 \boldsymbol{\beta} + 
 \begin{bmatrix}
  Z_1 & 0 \\
  0 & Z_2 
 \end{bmatrix}
\begin{bmatrix}
  g_1 \\
  g_2 
 \end{bmatrix}
  + \mathbf{e} \\
    & = 
 \begin{bmatrix}
  X_1 \\
  X_2 
 \end{bmatrix}
 \boldsymbol{\beta} + 
 \begin{bmatrix}
  Z_1 & 0 \\
  0 & Z_2 
 \end{bmatrix}
\begin{bmatrix}
  \mathbf{K}_{12}\mathbf{K}_{22}^{-1}\mathbf{W}_2\boldsymbol{\alpha} + \boldsymbol{\epsilon}  \\
  \mathbf{W}_2\boldsymbol{\alpha} \\
 \end{bmatrix}
  + \mathbf{e} \\
    & = 
 \begin{bmatrix}
  X_1 \\
  X_2 
 \end{bmatrix}
 \boldsymbol{\beta} + 
 \begin{bmatrix}
  Z_1 & 0 \\
  0 & Z_2 
 \end{bmatrix}
\begin{bmatrix}
  \hat{\mathbf{W}_1}\boldsymbol{\alpha} + \boldsymbol{\epsilon} \\
  \mathbf{W}_2\boldsymbol{\alpha} \\
 \end{bmatrix}
  + \mathbf{e} \\
\end{align}

@@

\begin{align} \label{eq:final-model}
\mathbf{y} &= \mathbf{Xb} +
\sum_{l=1}^{L} \mathbf{Z}_{l}\mathbf{W}_{l} \boldsymbol{\alpha}_{l} + 
\sum_{l=1}^{L} \mathbf{Z}_{l}\mathbf{U}_{l} \boldsymbol{\epsilon}_{l} +
\mathbf{e}.
\end{align}

>>>>

% Please add the following required packages to your document preamble:
% \usepackage{booktabs}
\begin{table}[]
\centering
\caption{Acronyms for the trait names}
\label{table:TraitKey}
\begin{tabular}{@{}llll@{}}
\toprule
Trait              & Abbreviation & Data Set  & Experiment \\ \midrule
Cob weight         & CW           & Diversity & Exp1       \\
Days to silking    & DS           & Diversity & Exp1       \\
Ear diameter       & ED           & Diversity & Exp1       \\
100-grain weight   & GW           & Diversity & Exp1       \\
Kernel width       & KW           & Diversity & Exp1       \\
Plant height       & PH           & Diversity & Exp1       \\
Dry matter yield   & DMY          & Hybrid    & Exp2       \\
Dry matter content & DMC          & Hybrid    & Exp2       \\
Lignin             & ADL          & Hybrid    & Exp2       \\
Fat                & FAT          & Hybrid    & Exp2       \\
Protein            & PRO          & Hybrid    & Exp2       \\
Starch             & STA          & Hybrid    & Exp2       \\
Sugar              & SUG          & Hybrid    & Exp2       \\ \bottomrule
\end{tabular}
\end{table}

@@

\Genetics2level{Prediction}
\subsection{Kernels}
BLUP models are computationally attractive when the number of features $p$
exceeds the number of genotypes $n$ and are equivalent to a selection index
when fixed effects have been accomodated by the dependent variable 
\cite{Mrode2014}.
Depending on the data set, up to three predictors were available for agronomic
trait predictions, namely pedigree data (P), genomic data (G) and
transcriptompic data (T).
The corresponding feature matrix of the inbred lines ($\mathbf{M}$) has
dimensions $n \times p$ where $n$ pertains to the number of genotypes.
Note that, for the sake of clarity, we do not distinguish between models for
Exp1 and Exp2 at this point.
In the case of Exp2 the term `genotype' refers strictly to the parent lines.
In generating kernels for each predictor, all features in $\mathbf{M}$ were
centered and standardized to unit variance.
Then, each kernel can be defined as

\begin{equation} \label{eq:GenomicRelationship}
  \mathbf{K} = \frac{1}{M} \mathbf{M} \mathbf{M}^{\top},
\end{equation}

where $M$ denotes the number of features~\cite{VanRaden2008}.
In the case of pedigree data (P), coancestry coefficients were used directly
for $\mathbf{K}$.

\subsection{Genetic model}
The general model for genetic values was:

\begin{equation} \label{eq:KBLUPModel}
  \mathbf{y} = \mu + \mathbf{Z} \mathbf{g} + \mathbf{e},
\end{equation}

where $\mathbf{y}$ is the vector of observed inbred line or hybrid performance,
respectively, $\mu$ is the fixed model intercept, and $\mathbf{e}$ is the 
residual error.
For the inbred lines we fit random genotype effects $\mathbf{g}$.
For both experiments, genetic effects $\mathbf{g}$ are associated with the
vector of phenotypic performance $\mathbf{y}$ through a corresponding design
matrix $\mathbf{Z}$.
The random effects ($\mathbf{g}$) have expectation zero and variance equal to
$\mathbf{K} \sigma^{2}_{g}$ and $\mathbf{I} \sigma^2_{e}$ for the residual
error.
Here, we did not consider the inclusion of dominance effects for the hybrid
material after a previous investigation into the predictive ability of various
``omics'' predictors in the same material concluded no considerable gains 
\cite{Westhues2017}.
All prediction models were implemented using the R-packages \textit{BGLR}
\cite{Perez2014} and \textit{sspredr}
\url{https://github.com/mwesthues/sspredr}.

\subsection{Single step prediction}
If the genetic values are fully captured by the features in $\mathbf{M}$,
Eq.~\ref{eq:KBLUPModel} can then be expressed as

\begin{align} \label{eq:KBLUPModelSimplified}
	\mathbf{y} &= \mu + \mathbf{Z} \mathbf{g} + \boldsymbol{e} \\
	&= \mu + \mathbf{Z} \mathbf{M} \boldsymbol{\alpha} + \boldsymbol{e},
\end{align}

\cite{Fernando2014} because the ridge regression model is equivalent to the
BLUP model~\cite{Ruppert2003}.
In Eq.~\ref{eq:KBLUPModelSimplified}, $\mathbf{M}$ is a matrix of feature
covariates and $\boldsymbol{\alpha}$ is the vector of random, partial-regression
coefficients of the feature covariates.

Hence, the breeding values in a prediction model are given by:
\begin{equation} \label{eq:mrnaebv}
	\mathbf{\hat{g}} = \mathbf{M}\boldsymbol{\hat{\alpha}},
\end{equation}

where $\boldsymbol{\hat{\alpha}}$ is the solution to the ridge regression model
(Eq.~\ref{eq:KBLUPModel}).

Consider now the situation in which one predictor is complete in the sense that
the whole population is equipped with features for that predictor whereas only
a subset of that population has information on a second predictor.
We can take a similar route as in \citeNP{Fernando2014} and impute covariates 
of the incomplete predictor by using covariates of the complete predictor.
Let the subscript $1$ denote individuals covered only by the complete
predictor whereas individuals covered by both, the complete and the
incomplete predictor, are indicated by the subscript $2$.
Further, we center the covariates of the incomplete predictor and denote it as
$\mathbf{W}$.
Correspondingly, $\mathbf{K}$ corresponds to the kernel for the complete
predictor.
In this case we need to partition the vector of genetic values $\mathbf{g}$
into a component $\mathbf{g}_{1}$ (\textit{i.e.,} genetic values of genotypes
covered only by the complete predictor) and a component
$\mathbf{g}_{2} = \mathbf{W}_{2}\boldsymbol{\alpha}$ (\textit{i.e.,} genetic
values of genotypes covered by both predictors, where $\boldsymbol{\alpha}$
pertains to the random feature effects of the incomplete predictor).
Under multivariate normality, the vector $\mathbf{g}_1$ can be written as the
sum of the conditional expectation given $\mathbf{g}_2$ and a residual error
term:
\begin{align} \label{eq:mrna1}
	\mathbf{g_1} &= E(\mathbf{g}_1|\mathbf{g}_2) + \boldsymbol{\epsilon} \\
	&= \mathbf{K}_{12}\mathbf{K}^{-1}_{22}\mathbf{W}_{2}
	\boldsymbol{\hat{\alpha}} + (\mathbf{g_1} - \mathbf{K}_{12}
	\mathbf{K}^{-1}_{2}
	\boldsymbol{\hat{\alpha}}) \\
	&= \mathbf{\hat{g}}_1 + \boldsymbol{\epsilon},
\end{align}

where $\mathbf{K}_{ij}$ are partitions of $\mathbf{K}$ that correspond to 
$\mathbf{g}_{1}$ and $\mathbf{g}_{2}$, respectively~\cite{Fernando2014} and
where $\mathbf{\hat{g}}_{1}$ pertains to the imputed genetic values in the
incomplete predictor.

The covariance matrix of $\boldsymbol{\epsilon}$ is $\mathbf{K}_{11} -
\mathbf{K}_{12}\mathbf{K}^{-1}_{22}\mathbf{K}_{21} =
(\mathbf{K}^{11})^{-1}$
\cite{Legarra2009}.

@@

\begin{equation} \label{eq:single-step-submatrices}
\mathbf{X} =
\begin{bmatrix}
  \mathbf{X}_1 \\
  \mathbf{X}_2
 \end{bmatrix},
 \mathbf{W} =
\begin{bmatrix}
  \mathbf{Z}_1\mathbf{\hat{W}}_{1} \\
  \mathbf{W}_2
 \end{bmatrix},
 \mathbf{U} =
\begin{bmatrix}
  \mathbf{Z}_1 \\
  0
 \end{bmatrix},
\end{equation}

where $\mathbf{\hat{W}}_{1} = \mathbf{K}_{12} 
\mathbf{K}^{-1}_{22} \mathbf{W}_{2}$.

@@

\begin{align} \label{eq:entangled-augmented-single-step-model}
\begin{bmatrix}
  \mathbf{y}_1 \\
  \mathbf{y}_2
 \end{bmatrix}
& =
 \begin{bmatrix}
  \mathbf{X}_1 \\
  \mathbf{X}_2
 \end{bmatrix}
 \boldsymbol{\beta} +
 \begin{bmatrix}
  \mathbf{Z}_1 & \boldsymbol{0} \\
  \boldsymbol{0} & \mathbf{Z}_2
 \end{bmatrix}
\begin{bmatrix}
  \mathbf{g}_1 \\
  \mathbf{g}_2
 \end{bmatrix}
  + \mathbf{e} \\
    & =
 \begin{bmatrix}
 \mathbf{X}_1 \\
 \mathbf{X}_2
 \end{bmatrix}
 \boldsymbol{\beta} +
 \begin{bmatrix}
  \mathbf{Z}_1 & \boldsymbol{0} \\
  \boldsymbol{0} & \mathbf{Z}_2
 \end{bmatrix}
\begin{bmatrix}
  \mathbf{K}_{12}\mathbf{K}^{-1}_{22}\mathbf{W}_2
  \boldsymbol{\alpha} + \boldsymbol{\epsilon}  \\
  \mathbf{W}_2\boldsymbol{\alpha} \\
 \end{bmatrix}
  + \mathbf{e} \\
    & =
 \begin{bmatrix}
  \mathbf{X}_1 \\
  \mathbf{X}_2
 \end{bmatrix}
 \boldsymbol{\beta} +
 \begin{bmatrix}
  \mathbf{Z}_1 & \boldsymbol{0} \\
  \boldsymbol{0} & \mathbf{Z}_2
 \end{bmatrix}
\begin{bmatrix}
  \mathbf{\hat{W}}_{1}\boldsymbol{\alpha} + \boldsymbol{\epsilon} \\
  \mathbf{W}_2\boldsymbol{\alpha} \\
 \end{bmatrix}
  + \mathbf{e} \\
\end{align}

@@

\begin{align} \label{eq:final-model}
\mathbf{y} &= \mathbf{Xb} +
\mathbf{Z}\mathbf{W} \boldsymbol{\alpha}_{l} + 
\sum_{l=1}^{L} \mathbf{Z}_{l}\mathbf{U}_{l} \boldsymbol{\epsilon}_{l} +
\mathbf{e}.
\end{align}
